WORKING TITLE: The adaptive value of green nest material across birds: A systematic review and meta-analysis

### 2025-07-08 docs/manuscript_GNM.qmd
Nothing much changed.. just pushing from desktop to now continue working from my laptop

First, the model using lnRR showed an average increase in fitness of `{r} round((exp(as.double(coef(intercept_lnRR)))-1)*100,1)`% in nests with an experimental increase in GNM \[@fig-mean-effect (A); lnRR = `{r} sprintf("%.3f", coef(intercept_lnRR))`, 95% CI: \[`{r} sprintf("%.3f", intercept_lnRR$ci.lb)` to `{r} sprintf("%.3f", intercept_lnRR$ci.ub)`\], 95% PI: \[`{r} sprintf("%.3f", predict(intercept_lnRR)$pi.lb)` to `{r} sprintf("%.3f", predict(intercept_lnRR)$pi.ub)`\], p-value = `{r} sprintf("%.3f", intercept_lnRR$pval)`, k = `{r} intercept_lnRR$k`, n = `{r} length(intercept_lnRR$s.levels.f[[1]])`\] but heterogeneity among effect sizes was high. Total raw heterogeneity (𝜎^2^) was `{r} sprintf("%.3f",sum(intercept_lnRR$sigma2))` (Q = `{r} sprintf("%.1f",sum(intercept_lnRR$QE))`, p \< 0.001), the measures of relative (I^2^~total~ = `{r} sprintf("%.1f",orchaRd::i2_ml(intercept_lnRR)[1])`%) and standardised heterogeneity (CVH2~total~ = `{r} sprintf("%.2f", orchaRd::cvh2_ml(intercept_lnRR)[1])` and M2~total =~ `{r} sprintf("%.2f", orchaRd::m2_ml(intercept_lnRR)[1])`) suggested high variability across effect sizes. Second, the model using SMD(H) showed a small positive and statistically significant effect of an experimental increase of GNM on fitness \[@fig-mean-effect (B); SMDH = `{r} sprintf("%.3f", coef(intercept_SMDH))`, 95% CI: \[`{r} sprintf("%.3f",(intercept_SMDH)$ci.lb)` to `{r} sprintf("%.3f",(intercept_SMDH)$ci.ub)`\], 95% PI: \[`{r} sprintf("%.3f", predict(intercept_SMDH)$pi.lb)` to `{r} sprintf("%.3f", predict(intercept_SMDH)$pi.ub)`\], p-value = `{r} sprintf("%.3f",intercept_SMDH$pval)`, k = `{r} intercept_SMDH$k`, n = `{r} length(intercept_SMDH$s.levels.f[[1]])`\], but heterogeneity among effect sizes was moderate to high. Total raw heterogeneity (𝜎^2^) was ``` {r}``sprintf("%.3f",sum(intercept_SMDH$sigma2)) ``` (Q = `{r} sprintf("%.1f",sum(intercept_SMDH$QE))`, p \< 0.001), relative heterogeneity was substantially smaller compared to that of lnRR (I^2^~total~ = `{r} sprintf("%.1f",orchaRd::i2_ml(intercept_SMDH)[1])`%) and the standardised heterogeneity metrics (CVH2~total~ = `{r} sprintf("%.2f", orchaRd::cvh2_ml(intercept_SMDH)[1])` and M2~total =~ `{r} sprintf("%.2f",orchaRd::m2_ml(intercept_SMDH)[1])`) lied between the 50th and 75th percentile of empirically derived estimates from meta-analyses in ecology and evolution [@yang2024a]. For lnRR, these heterogeneity metrics exceeded the 75th percentile benchmark (see @supp-tab-heterogeneity for detailed estimates of heterogeneity).

@@

Third, there is a positive statistically non-significant effect of increased GNM on fitness proxies relating to pathogen and parasitic load in case of arthropods and a negative non-significant effect in case of micro-organisms. \[[\@figsupp-MR](A-B); \[lnRR~arthropod~ = `{r} summary_lnRR_parasite$estimate[1]`, CI: `{r} summary_lnRR_parasite$CI[1]`, PI: `{r} summary_lnRR_parasite$PI[1]`, p-value = `{r} summary_lnRR_parasite$p_value[1]`, k = `{r} summary_lnRR_parasite$k[1]`, n = `{r} summary_lnRR_parasite$n[1]` and SMDH~arthropod~ = `{r} summary_SMDH_parasite$estimate[1]`, CI: `{r} summary_SMDH_parasite$CI[1]`, PI: `{r} summary_SMDH_parasite$PI[1]`, p-value = `{r} summary_SMDH_parasite$p_value[1]`, k = `{r} summary_SMDH_parasite$k[1]`, n = `{r} summary_SMDH_parasite$n[1]`; lnRR~mirco-organisms~ = `{r} summary_lnRR_parasite$estimate[2]`, CI: `{r} summary_lnRR_parasite$CI[2]`, PI: `{r} summary_lnRR_parasite$PI[2]`, p-value = `{r} summary_lnRR_parasite$p_value[2]`, k = `{r} summary_lnRR_parasite$k[2]`, n = `{r} summary_lnRR_parasite$n[2]` and SMDH~micro-organisms~ = `{r} summary_SMDH_parasite$estimate[2]`, CI: `{r} summary_SMDH_parasite$CI[2]`, PI: `{r} summary_SMDH_parasite$PI[2]`, p-value = `{r} summary_SMDH_parasite$p_value[2]`, k = `{r} summary_SMDH_parasite$k[2]`, n = `{r} summary_SMDH_parasite$n[2]`\]. The type of parasite or pathogen explained very little heterogeneity ($R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_parasite)*100,1))[1]`% and $R^2_{marginal(SMDH)}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_parasite)*100,1))[1]`%).

Most bird species, showed a small positive non-significant increase it fitness with an experimental increase in GNM, although some species showed negative but similarly non-significant effects \[[\@figsupp-MR](C-D); Table X\]. However, bird species identity explained a proportion of heterogeneity in the effect sizes ($R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_birds)*100,1))[1]`% and $R^2_{marginal(SMDH)}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_birds)*100,1))[1]`%).

Among different types of fitness trait, only morphological trait showed positive effect of GNM on fitness, which was statistically significant for SMDH \[lnRR~morphology~ = `{r} summary_lnRR_trait$estimate[2]`, CI: `{r} summary_lnRR_trait$CI[2]`, PI: `{r} summary_lnRR_trait$PI[2]`, p-value = `{r} summary_lnRR_trait$p_value[2]`, k = `{r} summary_lnRR_trait$k[2]`, n = `{r} summary_lnRR_trait$n[2]` and SMDH~morphology~ = `{r} summary_SMDH_trait$estimate[2]`, CI: `{r} summary_SMDH_trait$CI[2]`, PI: `{r} summary_SMDH_trait$PI[2]`, p-value = `{r} summary_SMDH_trait$p_value[2]`, k = `{r} summary_SMDH_trait$k[2]`, n = `{r} summary_SMDH_trait$n[2]`\]. Parasite and pathogen related fitness traits showed a small negative, non-significant effect in case of lnRR \[lnRR~parasite-pathogen~ = `{r} summary_lnRR_trait$estimate[3]`, CI: `{r} summary_lnRR_trait$CI[3]`, PI: `{r} summary_lnRR_trait$PI[3]`, p-value = `{r} summary_lnRR_trait$p_value[3]`, k = `{r} summary_lnRR_trait$k[3]`, n = `{r} summary_lnRR_trait$n[3]` and SMDH~parasite-pathogen~ = `{r} summary_SMDH_trait$estimate[3]`, CI: `{r} summary_SMDH_trait$CI[3]`, PI: `{r} summary_SMDH_trait$PI[3]`, p-value = `{r} summary_SMDH_trait$p_value[3]`, k = `{r} summary_SMDH_trait$k[3]`, n = `{r} summary_SMDH_trait$n[3]`\]. All other traits including behaviour, physiology, reproduction and phenology, showed small, non-significant effects [\@figsupp-MR](E-F); Table X\]. Trait type accounted for a small amount of heterogeneity ($R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_trait)*100,1))[1]`% and $R^2_{marginal(SMDH)}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_trait)*100,1))[1]`%).

@@

Summary of the results.. wrap up with a few points

@@

-   Discussing your results.. effect sizes

@@

Our systematic review and meta-analysis found a small positive increase in fitness benefits in the bird nests with increased GNM, though the strength of these effects depended on the effect size metric used. SMDH showed a larger effect with the There is high heterogeneity among the effect sizes, emphasizing the context-dependence of this benefit. Importantly, our results reveal that the commonly studied mechanistic hypotheses of GNM---whether through sexual signalling, parasite reduction, or direct nestling benefits---lack consistent empirical support across studies.

The While lnRR and SMDH differ in their measurement scale, both effect sizes suggest a small positive but non-conclusive effect of GNM on fitness. These differences in overall effect size magnitude likely reflect the underlying biological and statistical assumptions - lnRR incorporates the mean difference at a multiplicative scale whereas SMDH expresses the difference on an additive scale. The difference in relative heterogeneity (I^2^) between the two effect sizes aligns with broader trends observed in meta-analysis in ecology and evolution (@yang2024).

>>>>

First, the model using lnRR showed an average increase in fitness of `{r} round((exp(as.double(coef(intercept_lnRR)))-1)*100,1)`% in nests with an experimental increase in GNM \[@fig-mean-effect (A); lnRR = `{r} sprintf("%.3f", coef(intercept_lnRR))`, 95% CI: \[`{r} sprintf("%.3f", intercept_lnRR$ci.lb)` to `{r} sprintf("%.3f", intercept_lnRR$ci.ub)`\], 95% PI: \[`{r} sprintf("%.3f", predict(intercept_lnRR)$pi.lb)` to `{r} sprintf("%.3f", predict(intercept_lnRR)$pi.ub)`\], p-value = `{r} sprintf("%.3f", intercept_lnRR$pval)`, k = `{r} intercept_lnRR$k`, n = `{r} length(intercept_lnRR$s.levels.f[[1]])`\] but heterogeneity among effect sizes was high. Total raw heterogeneity (𝜎^2^) was `{r} sprintf("%.3f",sum(intercept_lnRR$sigma2))` (Q = `{r} sprintf("%.1f",sum(intercept_lnRR$QE))`, p \< 0.001), the measures of relative (I^2^~total~ = `{r} sprintf("%.1f",orchaRd::i2_ml(intercept_lnRR)[1])`%) and standardised heterogeneity (CVH2~total~ = `{r} sprintf("%.2f", orchaRd::cvh2_ml(intercept_lnRR)[1])` and M2~total =~ `{r} sprintf("%.2f", orchaRd::m2_ml(intercept_lnRR)[1])`) suggested high variability across effect sizes. Second, the model using SMD(H) showed a small positive and statistically significant effect of an experimental increase of GNM on fitness \[@fig-mean-effect (B); SMD(H) = `{r} sprintf("%.3f", coef(intercept_SMDH))`, 95% CI: \[`{r} sprintf("%.3f",(intercept_SMDH)$ci.lb)` to `{r} sprintf("%.3f",(intercept_SMDH)$ci.ub)`\], 95% PI: \[`{r} sprintf("%.3f", predict(intercept_SMDH)$pi.lb)` to `{r} sprintf("%.3f", predict(intercept_SMDH)$pi.ub)`\], p-value = `{r} sprintf("%.3f",intercept_SMDH$pval)`, k = `{r} intercept_SMDH$k`, n = `{r} length(intercept_SMDH$s.levels.f[[1]])`\], but heterogeneity among effect sizes was moderate to high. Total raw heterogeneity (𝜎^2^) was `{r} sprintf("%.3f",sum(intercept_SMDH$sigma2))` (Q = `{r} sprintf("%.1f",sum(intercept_SMDH$QE))`, p \< 0.001), relative heterogeneity was substantially smaller compared to that of lnRR (I^2^~total~ = `{r} sprintf("%.1f",orchaRd::i2_ml(intercept_SMDH)[1])`%) and the standardised heterogeneity metrics (CVH2~total~ = `{r} sprintf("%.2f", orchaRd::cvh2_ml(intercept_SMDH)[1])` and M2~total =~ `{r} sprintf("%.2f",orchaRd::m2_ml(intercept_SMDH)[1])`) lied between the 50th and 75th percentile of empirically derived estimates from meta-analyses in ecology and evolution [@yang2024a]. For lnRR, these heterogeneity metrics exceeded the 75th percentile benchmark (see @supp-tab-heterogeneity for detailed estimates of heterogeneity).

@@

Third, there was a positive statistically non-significant effect of increased GNM on fitness proxies relating to pathogen and parasitic load in case of arthropods and a negative non-significant effect in case of micro-organisms. \[[\@figsupp-MR](A-B); \[lnRR~arthropod~ = `{r} summary_lnRR_parasite$estimate[1]`, CI: `{r} summary_lnRR_parasite$CI[1]`, PI: `{r} summary_lnRR_parasite$PI[1]`, p-value = `{r} summary_lnRR_parasite$p_value[1]`, k = `{r} summary_lnRR_parasite$k[1]`, n = `{r} summary_lnRR_parasite$n[1]` and SMDH~arthropod~ = `{r} summary_SMDH_parasite$estimate[1]`, CI: `{r} summary_SMDH_parasite$CI[1]`, PI: `{r} summary_SMDH_parasite$PI[1]`, p-value = `{r} summary_SMDH_parasite$p_value[1]`, k = `{r} summary_SMDH_parasite$k[1]`, n = `{r} summary_SMDH_parasite$n[1]`; lnRR~mirco-organisms~ = `{r} summary_lnRR_parasite$estimate[2]`, CI: `{r} summary_lnRR_parasite$CI[2]`, PI: `{r} summary_lnRR_parasite$PI[2]`, p-value = `{r} summary_lnRR_parasite$p_value[2]`, k = `{r} summary_lnRR_parasite$k[2]`, n = `{r} summary_lnRR_parasite$n[2]` and SMDH~micro-organisms~ = `{r} summary_SMDH_parasite$estimate[2]`, CI: `{r} summary_SMDH_parasite$CI[2]`, PI: `{r} summary_SMDH_parasite$PI[2]`, p-value = `{r} summary_SMDH_parasite$p_value[2]`, k = `{r} summary_SMDH_parasite$k[2]`, n = `{r} summary_SMDH_parasite$n[2]`\]. The type of parasite or pathogen explained very little heterogeneity ($R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_parasite)*100,1))[1]`% and $R^2_{marginal(SMDH)}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_parasite)*100,1))[1]`%).

Fourth, the bird species had a small positive statistically non-significant increase with an experimental increase in GNM, although some species showed negative but similarly non-significant effects \[[\@figsupp-MR](C-D); Table X\]. However, bird species identity explained a proportion of heterogeneity in the effect sizes ($R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_birds)*100,1))[1]`% and $R^2_{marginal(SMDH)}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_birds)*100,1))[1]`%).

Lastly, among different types of fitness trait, only morphological trait had a positive effect of GNM on fitness, which was statistically significant for SMD(H) \[lnRR~morphology~ = `{r} summary_lnRR_trait$estimate[2]`, CI: `{r} summary_lnRR_trait$CI[2]`, PI: `{r} summary_lnRR_trait$PI[2]`, p-value = `{r} summary_lnRR_trait$p_value[2]`, k = `{r} summary_lnRR_trait$k[2]`, n = `{r} summary_lnRR_trait$n[2]` and SMDH~morphology~ = `{r} summary_SMDH_trait$estimate[2]`, CI: `{r} summary_SMDH_trait$CI[2]`, PI: `{r} summary_SMDH_trait$PI[2]`, p-value = `{r} summary_SMDH_trait$p_value[2]`, k = `{r} summary_SMDH_trait$k[2]`, n = `{r} summary_SMDH_trait$n[2]`\]. Parasite and pathogen related fitness traits showed a small negative, non-significant effect in case of lnRR \[lnRR~parasite-pathogen~ = `{r} summary_lnRR_trait$estimate[3]`, CI: `{r} summary_lnRR_trait$CI[3]`, PI: `{r} summary_lnRR_trait$PI[3]`, p-value = `{r} summary_lnRR_trait$p_value[3]`, k = `{r} summary_lnRR_trait$k[3]`, n = `{r} summary_lnRR_trait$n[3]` and SMDH~parasite-pathogen~ = `{r} summary_SMDH_trait$estimate[3]`, CI: `{r} summary_SMDH_trait$CI[3]`, PI: `{r} summary_SMDH_trait$PI[3]`, p-value = `{r} summary_SMDH_trait$p_value[3]`, k = `{r} summary_SMDH_trait$k[3]`, n = `{r} summary_SMDH_trait$n[3]`\]. All other traits including behaviour, physiology, reproduction and phenology, showed small, non-significant effects [\@figsupp-MR](E-F); Table X\]. Trait type accounted for some of the heterogeneity ($R^2_{marginal(lnRR)}$= `{r} as.double(round(orchaRd::r2_ml(lnRR_trait)*100,1))[1]`% and $R^2_{marginal(SMDH)}$= `{r} as.double(round(orchaRd::r2_ml(SMDH_trait)*100,1))[1]`%).

@@

Our pre-registered systematic review and meta-analysis provides the first quantitative synthesis of experimental studies on green nest material (GNM) across birds and reveals that, on average, supplementing nests with GNM yields only small benefits. The strength of the overall effects of GNM differed between the effect size metric used. Using the lnRR we found a non-significant 1.6% increase in overall fitness proxies, whereas the SMD(H) indicated a small but statistically significant positive effect (SMD(H)= 0.179). These differences in overall effect size magnitude likely reflect the underlying biological and statistical assumptions - lnRR incorporates the mean difference at a multiplicative scale whereas SMDH expresses the difference on an additive scale. There was moderate to high heterogeneity among the effect sizes, which we tried to understand using several biological and methodological moderators. The difference in relative heterogeneity (I^2^) between the two effect sizes aligns with broader trends observed in meta-analysis in ecology and evolution [@yang2024].

Contrary to long-standing verbal models, neither the courtship hypothesis (visual/olfactory signalling) nor the parental-care hypothesis (parasite repellency or pharmacological “drug” effects) emerged as a consistent driver of these small gains. Effect estimates associated with courtship traits were positive but non-significant, and those associated with nestling protection were essentially zero. In other words, current evidence does not allow us to attribute the weak average benefit of GNM to a single prevailing mechanism. This result mirrors the idiosyncratic outcomes scattered through the primary literature and indicates that future tests will need to measure multiple pathways simultaneously—ideally pairing behavioural observations with direct assays of ectoparasites, microbial load, and nest-air volatiles.

Importantly, our results reveal that the commonly studied mechanistic hypotheses of GNM --whether through sexual signalling, parasite reduction, or direct nestling benefits-- lack consistent empirical support across studies.

@@

@@

Temp Discussion:

**Discussion**

[manuscript_GNM]{.underline}. The discrepancy between metrics underscores how the inferred magnitude of benefit depends on the underlying scale: lnRR captures proportional shifts, whereas SMDᴴ reflects absolute shifts that are sensitive to among-study variance. Nonetheless, both converge on the same biological interpretation—if GNM is adaptive, the advantage appears subtle rather than dramatic, and it is expressed against a backdrop of substantial heterogeneity (I² ≈ 66–99 %).

Methodological context explained far more of the variance than biological hypothesis. Designs that compared nests with **no added material** versus **aromatic material** produced the largest positive effects (lnRR ≈ 0.15; SMDᴴ ≈ 0.84), whereas contrasts between **no material** and **non-aromatic plant matter** were negative, and the canonical **aromatic vs. non-aromatic** comparison was indistinguishable from zero. These patterns emphasise that simply adding any material can alter micro-climate, and that the perceived “aromatic benefit” may in part reflect removing a structural disadvantage in controls rather than an odour-specific mechanism. In addition, time of addition mattered: experiments that replenished GNM continuously throughout nesting detected stronger benefits (significant under SMDᴴ) than those adding it only before or after hatching, hinting that persistence of volatile release or parental display may be important [manuscript_GNM]{.underline}.

Despite these insights, most of the heterogeneity remains unexplained. Experimental design alone accounted for roughly a third of total variance, whereas parasite type, trait category, timing, and species identity each explained ≤ 8 % [manuscript_GNM]{.underline}. The residual scatter likely stems from taxonomic and geographic bias—82 % of effect sizes come from just three passerines (blue tit, spotless starling, European starling) studied at 16 sites, many of them by the same research groups. Publication-bias diagnostics were equivocal, yet small-study effects cannot be excluded given the prevalence of datasets with ≤ 5 nests per group and the tendency to publish novel positive outcomes. Together these limitations caution against broad generalisation and point to the need for coordinated multi-lab efforts on under-represented taxa (e.g., open-cup builders, raptors).

Several practical recommendations follow. First, future experiments should adopt fully factorial designs that separate **structure** (any plant vs. none) from **chemistry** (aromatic vs. non-aromatic) and track volatilome dynamics across the nesting cycle. Second, fitness proxies should move beyond morphology—currently the only trait category with a significant positive response—to harder endpoints such as fledging success, juvenile recruitment, and parental survival. Third, sharing raw data and negative results will be essential to reduce residual heterogeneity and clarify whether the modest average benefit we detect is biologically meaningful or a statistical artefact of design choices.

In summary, our review shows that adding green plant material to bird nests can increase fitness‐related measures, but the effect is small, context-dependent, and not yet pinned to a single mechanism. By refining experimental designs, broadening taxonomic coverage, and integrating chemical ecology with behavioural ecology, the next generation of studies can resolve whether GNM is truly an adaptive extended phenotype or merely a conspicuous but weak contributor to avian reproductive success.

Nicely looking colours!!!! Alright, alright. In short, ask me more if it is not clear. When you ran these 2 meta-regressions specifically, you normally would run them twice, once without the intercept from which you obtain the values you are presenting (i.e., the real lnRR and SMD(H) estimates specific for each off the three levels: Parental Care, Courtship, and Both), and then you would run them with the intercept. In the latter, the intercept, unless you changed the order of the levels in the moderator (which is a factor), would represent "Both", and the output would not tell you the difference between the estimate of level "Courtship" and "Both", and "Parental Care" and "Both" (i.e., the estimate is a difference, e.g. lnRRboth = 0.025 - lnRRcourtship = 0.038 = -0.13) and whether that difference is statistically significant. We want to show whether any differences between the levels are statistically significant, so we should present these p-values to say whether each of the pair-wise comparisons differs statistically or not. For you to know if "Courtship" and "Parental Care", you need to run an extra line of code, because from the model you only get the information about the difference between all levels of the factor (in this case, "Courtship" and "Parental Care") and the reference level AKA intercept (in this case "Both"). \
\
You can use:\
car::linearHypothesis(model, rbind(c(0,1,-1)))\
\
For more: [https://www.metafor-project.org/doku.php/tips:testing_factors_lincoms](https://www.google.com/url?q=https://www.metafor-project.org/doku.php/tips:testing_factors_lincoms&sa=D&source=docs&ust=1751963317237496&usg=AOvVaw3NHnLsaKPtM6xq0VTTbXZO)

Importantly, if the moderator would be a continuous moderator, e.g. publication year, you do not need to do any extra test, but should present the estimates for the intercept in the model as well as the estimates for the slope. When the moderator is continuous, you do not remove the intercept (i.e., no -1). Let me know if something is unclear. Good job, looking forward to reading the manuscript.
